Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-15 main.Rmd
calendar algebra minor changes

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple** and (2) **multiple-to-single** which entails the representation of multiple-order-up cyclic granularities from single order-up cyclic granularities and vice versa. The hierarchical structure of time creates a natural nested ordering which can produce _single-order-up_ or _multiple-order-up_ granularities. We shall use the notion of a hierarchy table and order to define them.

**Order** of a linear granularity can be comprehended as the level of coarseness associated with a linear granularity. For example, for two linear granularities $G$ and $H$, if $G$ groups into $H$ or $H$ is composed of higher number of granules of the bottom granularity than $G$, then $G$ is of lower order than $H$. In any hierarchy table, linear granularities are arranged from lowest to highest order.

**Hierarchy table** Let $H_n: (G, C, K)$ be a hierarchy table containing $n$ linear granularities. $G_{l}$ and $G_{m}$ represent the linear granularity of order $l$ and $m$ respectively with $l<m$. $K \equiv P(l,m)$ represents the period length of $G_{l}$ with respect to $G_{m}$ for circular granularity $C_{G(l),G(m)}$ and $K \equiv k(l,m)$ represents the operation to obtain $G_{m}$ from $G_{l}$ for quasi-circular granularity $C_{G(l),G(m)}$, $\forall l,m \in {1, 2, \dots,n}$. We assume that the cyclic granularity $C_{G(l),G(m)}$ is either circular or quasi-circular. It is difficult to position an aperiodic linear granularity in a hierarchy and hence the possibility of an aperiodic cyclic granularity is not considered.

@@

<!-- some discussions on primary units of calendar -->
**Example:** So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy table can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun. \autoref{tab:tab-mayan} represents the hierarchy table for the Mayan calendar.

@@

\begin{equation} \label{eq:eq7}
\begin{split}
C_{(G_l,G_m)}(z) & = L(C_{G_l,G_{l+1}}(z) + P(l, l+1)(C_{(G_{l+1},G_m)}(z))) \\
  & =   L(C_{(G_l,G_{l+1}}(z) + P(l, l+1)[ C_{G_{l+1},G_{l+2}}(z) + P(l+1, l+2)( C_{G_{l+2}, G_m}(z))) \\
  & =  L(C_{G_l,G_{l+1}}(z) + P(l, l+1)(C_{G_{l+1},G_{l+2}}(z)) + P(l, l+1)P(l+1, l+2)(C_{G_{l+2}, G_{m}}))\\
  & =  L(C_{G_l,G_{l+1}}(z) + P(l, l+1)(C_{G_{l+1},G_{l+2}}(z)) + P(l, l+2)(C_{G_{l+2},G_{l+m}}(z)))\\
  &\vdots\\
  & = L(\sum_{i=0}^{m - l - 1} P(l, l+i)(C_{G_{l+i},G_{l+i+1}}(z)))\\
\end{split}
\end{equation}

, for $l < m + 1$ and $P(i, j) = 1, \quad \forall i = j \in \{0, 1, \dots, m-l-1\}$

**Example:** Let us use Equation \ref{eq:eq7} to compute the multiple-order-up granularity $C_{uinal,katun}$ for Mayan calendar.

\begin{equation} \label{eq8}
\begin{split}
C_{uinal, baktun}(z)  = & L(C_{uinal, tun}(z) + P(uinal, tun)C_{tun,katun}(z) + P(uinal, katun)C_{katun, baktun}(z)) \\
               = & L(\lfloor z/20\rfloor \mod 18 + 18*\lfloor z/20*18\rfloor \mod 20  \\
               &~~~ + 18*20\lfloor z/20*18*20\rfloor \mod 20) \\
\end{split}
\end{equation}

@@

<!-- For example, with $C_{l, m}(z) = f(C_{l,m'}(z), C_{m',m}(z))$, -->
- $C_{l,m'}(z)$ is circular and $C_{m',m}(z)$ is quasi-circular
\begin{equation} \label{eq:multifromsingle-quasi1}
\begin{split}
C_{G_l,G_{l+2}}(z) & = L(C_{G_{l},G_{m'}}(z) + P(l, m')(C_{G_{m'},G_{m}}(z))) \\
\end{split}
\end{equation}

- $C_{l,m'}(z)$ is quasi-circular and $C_{m',m}(z)$ is circular
\begin{equation} \label{eq:multifromsingle-quasi2}
\begin{split}
C_{G_l,G_{l+2}}(z) & = L(C_{G_{l},G_{m'}}(z) + \sum_{w=0}^{C_{m',m}(z) -1}(\vert T_{w} \vert))\\
\end{split}
\end{equation}

@@

 - $C_{hour, month}(z) = L(C_{hour, day}(z) + P(hour, day)*C_{day, month}(z))$  
 - $C_{day, year}(z) = L(C_{day,month}(z) + \sum_{w=0}^{C_{month, year}(z)-1}(\vert T_{w} \vert))$, where, $T_w$ is such that $month(w) = \bigcup_{z \in T_w}day(z)$.

@@

\begin{equation} \label{eq:all-circular-multiple}
C_{G_{l_1}, G_{m_1}}(z) = C_{G_{l_2}, G_{m_2}}(\lfloor z/k(l_2,l_1) \rfloor\mod P(m_1, m_2))
\end{equation}

**Example:** Considering the same example of Mayan Calendar, it is possible to compute the single-order-up granularity tun-of-katun  given the multiple-order-up granularity uinal-baktun using equation \ref{eq:all-circular-multiple}.

\begin{equation} \label{eq10}
\begin{split}
C_{tun, katun}(z) =  L(\lfloor C_{uinal, baktun}(z)/18\rfloor \mod 20)
\end{split}
\end{equation}

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. 
Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

>>>>

The cyclic calendar algebra consists of broadly two kinds of operations:
(1) **single-to-multiple** and (2) **multiple-to-single** which entails the representation of _multiple-order-up_ cyclic granularities from _single-order-up_ cyclic granularities and vice versa. The hierarchical structure of time creates a natural nested ordering which can produce single-order-up or multiple-order-up granularities. We shall use the notion of a hierarchy table and order to define them.

**Order** of a linear granularity can be comprehended as the level of coarseness associated with a linear granularity. For example, for two linear granularities $G$ and $H$, if $G$ groups into or finer than $H$ or $H$ is composed of higher number of granules of the bottom granularity than $G$, then $G$ is of lower order than $H$. In any hierarchy table, linear granularities are arranged from lowest to highest order.

**Hierarchy table** Let $H_n: (G, C, K)$ be a hierarchy table containing $n$ linear granularities. $G_{l}$ and $G_{m}$ represent the linear granularity of order $l$ and $m$ respectively with $l<m$. $K \equiv P(l,m)$ represents the period length of the grouping $(G_{l}, G_{m})$ for circular granularity and $K \equiv k(l,m)$ represents the operation to obtain $G_{m}$ from $G_{l}$ for quasi-circular granularity $C_{G(l),G(m)}$, $\forall l,m \in {1, 2, \dots,n}$. We assume that the cyclic granularity $C_{G(l),G(m)}$ is either circular or quasi-circular. It is difficult to position an aperiodic linear granularity in a hierarchy and hence the possibility of an aperiodic cyclic granularity is not considered.

@@

<!-- some discussions on primary units of calendar -->
<!-- **Example:**  -->

So far we have used example of cyclic granularities from Gregorian calendar as it is the most widely used calendar. But it is far from being the only one. <!--All calendars fall under three types - solar, lunar or lunisolar/solilunar-->The day is the basic unit of time underlying most calendars [@Reingold2001-kf]. Various calendars, however, use different conventions to structure days into larger units: weeks, months, years and cycle of years. The French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds". Nevertheless, for any calendar a hierarchy table can be defined. For example, in Mayan calendar, one day was referred to as 1 kin and the calendar was structured such that 1 kin = 1 day; 1 uinal = 20 kin; 1 tun  = 18 uinal; 1 katun = 20 tun and 1 baktun = 20 katun. \autoref{tab:tab-mayan} represents the hierarchy table for the Mayan calendar.

@@

\begin{equation} \label{eq:eq7}
\begin{split}
C_{G_l,G_m}(z) & = C_{G_l,G_{l+1}}(z) + P(l, l+1)(C_{G_{l+1},G_m}(z)) \\
  & = C_{(G_l,G_{l+1}}(z) + P(l, l+1)[ C_{G_{l+1},G_{l+2}}(z) + P(l+1, l+2)]( C_{G_{l+2}, G_m}(z)) \\
  & = C_{G_l,G_{l+1}}(z) + P(l, l+1)(C_{G_{l+1},G_{l+2}}(z)) + P(l, l+1)P(l+1, l+2)(C_{G_{l+2}, G_{m}})\\
  & = C_{G_l,G_{l+1}}(z) + P(l, l+1)(C_{G_{l+1},G_{l+2}}(z)) + P(l, l+2)(C_{G_{l+2},G_{l+m}}(z))\\
  &\vdots\\
  & = \sum_{i=0}^{m - l - 1} P(l, l+i)(C_{G_{l+i},G_{l+i+1}}(z))\\
\end{split}
\end{equation}

, for $l < m - 1$ and $P(i, j) = 1, \quad \forall i = j \in \{0, 1, \dots, m-l-1\}$

**Example:** Let us use Equation \ref{eq:eq7} to compute the multiple-order-up granularity $C_{uinal,katun}$ for Mayan calendar with the index set  $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$ mapping "kins" to a set of positive integers.

\begin{equation} \label{eq8}
\begin{split}
C_{uinal, baktun}(z)  = & C_{uinal, tun}(z) + P(uinal, tun)C_{tun,katun}(z) + P(uinal, katun)C_{katun, baktun}(z) \\
               = & \lfloor z/20\rfloor \mod 18 + 18*\lfloor z/20*18\rfloor \mod 20  \\
               &~~~ + 18*20\lfloor z/20*18*20\rfloor \mod 20 \\
\end{split}
\end{equation}

@@

<!-- For example, with $C_{l, m}(z) = f(C_{l,m'}(z), C_{m',m}(z))$, -->
- $C_{l,m'}(z)$ is circular and $C_{m',m}(z)$ is quasi-circular
\begin{equation} \label{eq:multifromsingle-quasi1}
\begin{split}
C_{G_l,G_{m}}(z) & = C_{G_{l},G_{m'}}(z) + P(l, m')(C_{G_{m'},G_{m}}(z)) \\
\end{split}
\end{equation}

- $C_{l,m'}(z)$ is quasi-circular and $C_{m',m}(z)$ is circular
\begin{equation} \label{eq:multifromsingle-quasi2}
\begin{split}
C_{G_l,G_{m}}(z) & = C_{G_{l},G_{m'}}(z) + \sum_{w=0}^{C_{m',m}(z) -1}(\vert T_{w} \vert)\\
\end{split}
\end{equation}

@@

 - $C_{hour, month}(z) = C_{hour, day}(z) + P(hour, day)*C_{day, month}(z)$  
 - $C_{day, year}(z) = C_{day,month}(z) + \sum_{w=0}^{C_{month, year}(z)-1}(\vert T_{w} \vert)$, where, $T_w$ is such that $month(w) = \bigcup_{z \in T_w}day(z)$

@@

\begin{equation} \label{eq:all-circular-multiple}
C_{G_{l_1}, G_{m_1}}(z) = \lfloor C_{G_{l_2}, G_{m_2}}(z)/P(l_2,l_1) \rfloor \mod P(l_1, m_1)
\end{equation}

<!-- **Example:**  -->
Considering the same example of Mayan Calendar, it is possible to compute the single-order-up granularity tun-of-katun  given the multiple-order-up granularity uinal-baktun using equation \ref{eq:all-circular-multiple}.

\begin{equation} \label{eq10}
\begin{split}
C_{tun, katun}(z) =  \lfloor C_{uinal, baktun}(z)/18\rfloor \mod 20
\end{split}
\end{equation}

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}, let $S_{ij}$ be a subset of the index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. 
Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->
